---
title: "Who Can Live in Chicago?"
subtitle: "Investigating housing affordability trends using 2020 census data"
author: "Ethan Jantz"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE)
library(kableExtra)
library(tidyverse)
library(tidycensus)
library(patchwork)

ref_vars <- load_variables(2015, "acs5")

acs_vars <- c(
  total_population = "B01001_001",
  white_population = "B02001_002",
  black_population = "B02001_003",
  amind_population = "B02001_004",
  asian_population = "B02001_005",
  api_population = "B02001_006",
  other_population = "B02001_007",
  two_population = "B02001_008",
  nothispanic_population = "B03001_002",
  hispanic_population = "B03001_003",
  total_households = "B25003_001",
  owner_households = "B25003_002",
  renter_households = "B25003_003",
  vacant_units = "B25004_001",
  vacant_units_forrent = "B25004_002",
  total_families = "B05010_001",
  belowpoverty_families = "B05010_002",
  belowpoverty_own_families1 = "B17019_004",
  belowpoverty_own_families2 = "B17019_008",
  belowpoverty_own_families3 = "B17019_011",
  belowpoverty_rent_families1 = "B17019_005",
  belowpoverty_rent_families2 = "B17019_009",
  belowpoverty_rent_families3 = "B17019_012",
  median_income = "B25119_001",
  median_owner_income = "B25119_002",
  median_renter_income = "B25119_003",
  median_gross_rent = "B25031_001",
  median_gross_2br_rent = "B25031_004",
  median_home_value = "B25107_001"
)

inflation_table <- blscrapeR::inflation_adjust(2020) %>%
  filter(year %in% c(2015, 2020))
inflation_adjust_value <- inflation_table %>% 
  filter(year == 2015) %>%
  pull(adj_value)

chi_2020 <- get_acs("place",
                    variables = acs_vars,
                    year = 2020, 
                    output = "wide",
                    cache_table = T) %>%
  mutate(year = 2020)

chi_2015 <- get_acs("place",
                    variables = acs_vars,
                    year = 2015, 
                    output = "wide",
                    cache_table = T) %>%
  mutate(year = 2015,
         across(ends_with(c("_income", "_value", "_rent")), ~ .x / inflation_adjust_value))

data <- bind_rows(chi_2020, chi_2015) %>%
  filter(NAME == "Chicago city, Illinois") %>%
  select(city = NAME, year, ends_with("E")) %>%
  rename_with(gsub, pattern = "E$", replacement = "") %>%
  mutate(grouped_other_population = amind_population + api_population + other_population,
         belowpoverty_own_families = belowpoverty_own_families1 + belowpoverty_own_families2 + belowpoverty_own_families3,
         belowpoverty_rent_families = belowpoverty_rent_families1 + belowpoverty_rent_families2 + belowpoverty_rent_families3)
```

# Overview

The release of the 2020 5-year American Community Survey data provides an opportunity to explore how Chicago and its neighborhoods have changed over the past decade. This blog is the first in a series exploring housing affordability trends in Chicago. This data provides new insights into trends that were observed in our 2006 report [Affordable Housing Conditions and Outlook in Chicago: An Early Warning for Intervention](http://voorheescenter.red.uic.edu/wp-content/uploads/sites/122/2017/10/2006-Affordable-housing-conditions-and-outlook-in-Chicago.pdf). Notably, housing affordability pressures have continued to disproportionately affect low-income households leading to out-migration as they seek more affordable places to live. Obversely (or, on the other side of the coin?) Chicago has seen a growth in median income, though median income for renters has continued to stay well below that of homeowners and the city as a whole. Rental vacancies have declined as well, while rents and home values have increased by almost 20% over 5 years. These changes likely contribute to the continued decline of Chicago's black population as housing affordability pressures hit low-income households of color the hardest. 

## Analysis

```{r population total}
data %>%
  select(Year = year, Population = total_population, Households = total_households) %>%
  pivot_longer(
    cols = c(Households, Population),
    names_to = " "
  ) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "value"
  ) %>%
  select(` `, `2015`, `2020`) %>%
  mutate(Change = `2020` - `2015`,
         across(c(`2015`, `2020`, Change), scales::comma)) %>%
  kable() %>%
  kable_styling(latex_options = "striped")
```

Chicago lost an estimated 18,187 residents between 2015 and 2020. Meanwhile, Chicago gained an estimated 45,000 households between 2015 and 2020, indicating a shrink in household size. We will examine this and other questions on household composition in later posts. 

Proportionally, Chicago has seen declines in groups identifying as white, black, hispanic, and other (a category that includes anyone identifying as neither white, black, asian, or as two or more races) while groups identifying as asian and two or more races saw increases. People identifying as two or more races saw the largest change in Chicago's population, growing from 2.4% in 2015 to 5.2% in 2020. 

```{r demographic}
data %>%
  mutate(
    white_pct = white_population / total_population,
    black_pct = black_population / total_population,
    asian_pct = asian_population / total_population,
    # api_pct = api_population / total_population,
    other_pct = (other_population + api_population) / total_population,
    two_pct = two_population / total_population,
    # other_pct = grouped_other_population / total_population,
    hispanic_pct = hispanic_population / total_population,
    nothispanic_pct = nothispanic_population / total_population,
    total_pct = white_pct + black_pct + other_pct,
    # across(ends_with("_population"), scales::comma),
    # across(ends_with("_pct"), scales::percent)
         ) %>%
  select(
    year, #total_population, 
    White_pct = white_pct, #white_population, 
    Black_pct = black_pct, #black_population, 
    Asian_pct = asian_pct, #asian_population, 
    # API_pct = api_pct, #api_population,
    `Two or more_pct` = two_pct, #two_population, 
    Other_pct = other_pct, #other_population
    Hispanic_pct = hispanic_pct, #hispanic_population, 
    # nothispanic_pct, #nothispanic_population 
  ) %>%
  pivot_longer(
    cols = c(White_pct:Hispanic_pct),
    names_to = "Race",
    names_pattern = "(.*)_pct"
  ) %>%
  pivot_wider(
    names_from = "year",
    values_from = "value"
  ) %>%
  mutate(Change = `2020` - `2015`,
         across(c(`2020`:Change), scales::percent, accuracy = .01)) %>% 
  select(`Race/Ethnicity` = Race, `2015`, `2020`, Change) %>%
  kable() %>%
  kable_styling(latex_options = "striped")
```

Chicago saw a modest increase in its share of homeowners and decrease in renter households. Median gross rent increased by 19% (\$189) and the median home value grew by 20% (\$44,700). The rental vacancy rate decreased from 5.9% to 4.8%. This indicates increasing pressure on low-income households as rental units become less available and less affordable.

```{r tenure}
data %>%
  mutate(
    renter_pct = renter_households / total_households,
    owner_pct = owner_households / total_households,
    rental_vacancy_rate = vacant_units_forrent / (vacant_units + renter_households),
    vacant_forrent_units = vacant_units_forrent / vacant_units,
    # total_households = scales::comma(total_households),
    across(ends_with(c("units", "forrent")), scales::comma),
    across(ends_with(c("_rent", "income", "value")), scales::dollar, prefix = " $ "),
    across(ends_with(c("_pct", "rate")), scales::percent)) %>%
  select(Year = year, `Percent Renter Households` = renter_pct, `Median Gross Rent` = median_gross_rent, `Median Home Value` = median_home_value, `Rental Vacancy Rate` = rental_vacancy_rate) %>% 
  pivot_longer(
    cols = c(`Percent Renter Households`:`Rental Vacancy Rate`),
    names_to = " "
  ) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "value"
  ) %>%
  select(` `, `2015`, `2020`) %>%
  kable(format = "html", escape = F) %>%
  kable_styling(latex_options = "striped") %>%
  footnote(general = "<i>Rent and home values in inflation adjusted 2020 dollars ($)</i>", general_title = "", escape = F)
```

 The median income for Chicago increased by 27% (\$13,575). Median income for renters increased by 31% (\$10,417) and owners at 22% ($16,711). 

```{r income}
data %>%
  select(Year = year, `Median Income` = median_income, `Median Income (Renters)` = median_renter_income, `Median Income (Homeowners)` = median_owner_income) %>%
  mutate(across(contains("Income"), scales::dollar, prefix = " $ ")) %>% 
  pivot_longer(
    cols = c(`Median Income`:`Median Income (Homeowners)`),
    names_to = " "
  ) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "value"
  ) %>%
  select(` `, `2015`, `2020`) %>%
  kable() %>%
  kable_styling(latex_options = "striped") %>%
  footnote(general = "<i>Rent and home values in inflation adjusted 2020 dollars ($)</i>", general_title = "", escape = F)
```

Chicago saw a decrease of 46,595 families (a household of related individuals) between 2015 and 2020. The share of those families living in poverty dropped by 8%, indicating an out-migration of low-income families from the city. The loss of low-income families was largely driven by renters at a loss of 25,086. Chicago's tightening rental market and rising rents are making the city increasingly unaffordable for low-income families. 

```{r poverty}
data %>%
  mutate(poverty_pct = belowpoverty_families / total_families,
         poverty_rent_pct = belowpoverty_rent_families / total_families,
         poverty_own_pct = belowpoverty_own_families / total_families,
         across(ends_with("families"), scales::comma),
         across(ends_with("pct"), scales::percent)) %>%
  select(Year = year, `Families` = total_families, `Families Below Poverty` = poverty_pct, `Percent Renter Families Below Poverty` = poverty_rent_pct, `Percent Homeowner Families Below Poverty` = poverty_own_pct) %>% 
  pivot_longer(
    cols = c(`Families`:`Percent Homeowner Families Below Poverty`),
    names_to = " "
  ) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "value"
  ) %>%
  select(` `, `2015`, `2020`) %>%
  kable() %>%
  kable_styling(latex_options = "striped")
```


<!-- ## Chicago Community Areas Crosswalk -->

<!-- [The Census Bureau provides an explanation of the 2010 to 2020 tract relationships here.](https://www2.census.gov/geo/pdfs/maps-data/data/rel2020/tract/explanation_tab20_tract20_tract10.pdf) This file is joined to the tract relationship file provided by the [City of Chicago Data Portal](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Census-Tracts-2010/5jrd-6zik) to identify which 2020 decennial tracts are related to the Chicago Community Areas (CCAs) that the 2010 decennial tracts aggregate to. We will be releasing this crosswalk file for use by other researchers in a later blog post where we explore change at the CCA level.  -->

<!-- ```{r crosswalk, include = FALSE} -->
<!-- chi_tracts_2010 <- read_csv("https://data.cityofchicago.org/resource/74p9-q2aq.csv", col_types = "cccccccccc") -->

<!-- cook_tracts_2010 <- tigris::tracts(state = "IL", county = "Cook", year = 2010)  -->

<!-- chi_tracts_2010 <- cook_tracts_2010 %>% -->
<!--   right_join(chi_tracts_2010, by = c("GEOID10" = "geoid10")) -->

<!-- cook_tracts_2020 <- tigris::tracts(state = "IL", county = "Cook", year = 2020) -->

<!-- rel_file <- read.delim("https://www2.census.gov/geo/docs/maps-data/data/rel2020/tract/tab20_tract20_tract10_st17.txt", sep = "|", colClasses = "character") -->

<!-- new_rel_file <- chi_tracts_2010 %>% -->
<!--   select(GEOID10, commarea) %>% -->
<!--   left_join(rel_file %>% -->
<!--                select(GEOID_TRACT_10, GEOID_TRACT_20),  -->
<!--              by = c("GEOID10" = "GEOID_TRACT_10"))  -->

<!-- chi_tracts_2020 <- new_rel_file %>% -->
<!--   sf::st_drop_geometry() %>% -->
<!--   left_join(cook_tracts_2020, -->
<!--             by = c("GEOID_TRACT_20" = "GEOID")) %>% -->
<!--   select(-GEOID10, GEOID = GEOID_TRACT_20, everything()) -->

<!-- p2010 <- chi_tracts_2010 %>% -->
<!--   group_by(commarea) %>% -->
<!--   summarize() %>% -->
<!--   ggplot(aes(geometry = geometry)) + -->
<!--   geom_sf(aes(fill = commarea)) + -->
<!--   theme(legend.position = "none") -->

<!-- p2020 <- cook_tracts_2020 %>% -->
<!--   select(GEOID, geometry) %>% -->
<!--   left_join(chi_tracts_2020, by = c("GEOID")) %>% -->
<!--   drop_na() %>% -->
<!--   group_by(commarea) %>% -->
<!--   summarize() %>% -->
<!--   ggplot(aes(geometry = geometry.x)) + -->
<!--   geom_sf(aes(fill = commarea)) + -->
<!--   theme(legend.position = "none") -->

<!-- p2010 + p2020 -->
<!-- ``` -->