---
title: "Chicago Rental Unit Count 2020"
author: "Ethan Jantz, UIC Voorhees Center"
date: "4/10/2022"
output: html_document
---

A common question asked when drafting policy or legislation related to renters in Chicago is "how many rental units would this affect?" In the case where exemptions are created for owner-occupied bulidings with 6 or fewer units this number can be difficult to pin down as data is not collected on this specific case. This analysis aims to put together a rough estimate using Census microdata from 2020. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(kableExtra)
library(tidyverse)

data <- ipumsr::read_ipums_micro(ddi = here::here("Data", "IPUMS", "chicago_units.xml"),
                                 data_file = here::here("Data", "IPUMS", "chicago_units.csv.gz"))
```

Census microdata is organized, harmonized, and held on the Integrated Public Use Microdata Series (IPUMS) database by the University of Minnesota. I utilize the `VACANCY`, `OWNERSHP`, and `UNITSSTR` variables to identify rental units in buildings covered by Chicago's Residential Landlord Tenant Ordinance. `UNITSSTR` is organized into the following categories:

`r ipumsr::ipums_val_labels(data, UNITSSTR) %>% kable() %>% kable_styling()`

The categories relevant to this question are categories 3, 4, 5, 6, and 7. Rental properties are identified by the `OWNERSHP` variable, which identifies owner and renter occupied units. `VACANCY` allows for the identification of vacant units. Unfortunately, owner-occupation is not a variable provided by the Census. Work has been done to identify owner-occupied units using data from the Cook County Assessor's Office, but this has not been performed on more recent data. Thus the number of owner-occupied units will be a subset of the number of units identified using this data. 

Category 7 in `UNITSSTR` idenifies units in 5 - 9 unit buildings. Since we are attempting to identify rental units in buildings composed of 6 or fewer units I will provide three numbers. The first will be the total sum of renter-occupied units in buildings with 9 or fewer units. The second will be the total sume of renter-occupied units in buildings with 4 or fewer units. The final number will be a count with a rough estimate of units in buildings with 6 or fewer units. Since we cannot identify the distribution of unit count within the 5 - 9 unit category we will pull a proportion of those units, 15%, to add to the count. The equation for this estimate is as follows:

> *Count of Units in 4 or fewer unit buildings* + (*Count of units in 5 - 9 unit buildings* * .15)

```{r }
tibble(
  Category = c("9 or Fewer", "4 or fewer", "6 or Fewer*"),
  Count = c(data %>%
              distinct(SERIAL, .keep_all = T) %>%
              filter(UNITSSTR %in% c(3, 4, 5, 6, 7),
                     OWNERSHP == 2,
                     VACANCY %in% c(0, 1)) %>%
              count(wt = HHWT) %>%
              pull(n) %>%
              scales::comma(), 
            data %>%
              distinct(SERIAL, .keep_all = T) %>%
              filter(UNITSSTR %in% c(3, 4, 5, 6),
                     OWNERSHP == 2,
                     VACANCY %in% c(0, 1)) %>%
              count(wt = HHWT) %>%
              pull(n) %>%
              scales::comma(),
            data %>%
              distinct(SERIAL, .keep_all = T) %>%
              filter(UNITSSTR %in% c(3, 4, 5, 6),
                     OWNERSHP == 2,
                     VACANCY %in% c(0, 1)) %>%
              count(wt = HHWT) %>%
              pull(n) %>%
              `+`(
                data %>%
              distinct(SERIAL, .keep_all = T) %>%
              filter(UNITSSTR %in% c(7),
                     OWNERSHP == 2,
                     VACANCY %in% c(0, 1)) %>%
              count(wt = HHWT) %>%
              pull(n) %>%
                `*`(.15)
              ) %>%
              scales::comma(accuracy = 1)
  )
) %>%
  kable() %>%
  kable_styling(latex_options = "striped") %>%
  footnote(general_title = "", general = "<i>* This is an estimated number</i>", escape = F)
```

The above table provides esimates of the total number of units occupied by renters. Not all of these units exist in owner-occupied buildings, and without any official administrative data providing a survey count of owner-occupied units in Chicago we can only find approximate numbers. A policy that would assist in identifying these properties would be a rental registry, a program that has already been implemented in Chicago for [short-term rentals such as AirBnB listings](https://www.chicago.gov/city/en/depts/bacp/supp_info/sharedhousingandaccomodationslicensing.html). Rental registries are already utilized in large cities such as [New York City](https://www1.nyc.gov/site/hpd/services-and-information/register-your-property.page) and [Los Angeles](https://housing.lacity.org/rental-property-owners/rent-registry). The implementation of a rental and landlord register in Chicago, as well as the publication of this register on the City's open data portal, would allow for a better understanding of the state of Chicago's housing stock. 
